Towards a classification framework for social machines

### 2013-02-22 latex/paper_new.tex
Polyarchy simplified and reduced in size.
Added refined list of constructs, removed sections and added outline of new section

To describe social machines we roughly rely on four different categories of classification features:

\begin{description}
  \item [The types of tasks undertaken by the social component and the resulting contributions]: social machines, just as any computing system, are built with a given purpose in mind, and exhibit functionality by which this purpose can be achieved. Users and the emerging social network interact with the system and with each other to carry out specific tasks by which they contribute to the 'computation'. This includes the creation and refinement of digital artifacts by executing given actions, possibly in collaboration with other participants. For example, in GalaxyZoo\footnote{\url{http://www.galaxyzoo.org}} people classify galaxies with collective performance as good as professional astronomers, while in reCaptcha\footnote{\url{http://www.google.com/recaptcha}} they refine the data that describes digital manuscripts. Sometimes, systems look for innovative ways of getting people to engage with the systems, including game narratives, like in EteRNA\footnote{\url{http://eterna.cmu.edu}}, in which  participants design RNA  that is evaluated automatically using simulators and then refined via human effort, or monetary rewards, like in Amazon's Mechanical Turk, a platform in which users undertake routine (micro) tasks such as object classification, audio transcription and text translation.

  \item [The roles of the participants and the social network they form]:

  \item [The motivating factors and incentive mechanisms]: system features influencing user behavior and the social network dynamics

  \item [Interaction between social and machine component]: how are human contributions used to compute things, including quality control, spam, workflow, aggregation of results
\end{description}

@@

\subsection{Methodology}

@@

\subsection{Classification dimensions}

\subsubsection{The polyarchical relationship of social machines}
While determining a set of social machines, one clear distinction was between social machine frameworks, such as MediaWiki and Ushahidi, which enable
social machines to be created, and instantiations of those frameworks into social machines that have been created for a specific purpose, such as Wikipedia
and Ushahidi Haiti. However, through further investigation, we discovered that the distinction between framework and instantiations was not the only
differentiator. Specifically, that some social machines could also have sub-machines or communities within them, that are working to solve different problems,
within the context of the larger machine. For example, Wikipedia is an instantiation of the MediaWiki framework, but there are a large number of
communities within Wikipedia that work on domain-specific areas, such as those contributing in areas of science and technology\footnote{Visualisation of Wikipedia Science Communities: \url{http://www.olihb.com/WikiCommunities/}}. However, we must also note that wikipedia is primarily split into communities of language, with the English, German and French wikipedias having the largest number of articles. It is therefore possible to analyse wikipedia into a hierarchy, from MediaWiki, to the languages of Wikipedia, down to the domain-specific communities. We explored this idea further, to include other social machines, and links where social machines utilise other social machines in order to function. For example, the social machine used by the Obama '12 US Presidential campaign~\cite{obamakieron}, which relied on social machines such as Twitter and Facebook.

@@

\begin{figure*}[htb]
\begin{center}
\includegraphics[width=18cm]{img/polyarchy.pdf}
\caption{A polyarchy of Social Machines, illustrating the infrastructure and frameworks used by social machines, and machine-machine usage.} \label{polyarchy}
\end{center}
\end{figure*}

@@

\subsubsection{Multi-faceted motivation, and who-gets-what benefit}

%We identified a hierarchy of constructs pertaining to the motivation people have for using social
%machines with top level constructs of:
%
%\begin{enumerate}
%\item Hedonistic
%\item Financial
%\item To be informed
%\item To help others
%\end{enumerate}
%
%The sub constructs are illustrated in Figure~\ref{motivation}, and we relate them to Maslow's hierarchy of needs~\cite{maslow}, (e.g., self-actualisation, security, respect of others), in various ways.
%
%\begin{figure*}[htb]
%\begin{center}
%\includegraphics[width=18cm]{img/motivation.pdf}
%\caption{Motivation hierarchy of participants using Social Machines, related to Maslow's hierarchy of needs.} \label{motivation}
%\end{center}
%\end{figure*}

The motivations of users can be identified and mapped to existing human needs and motivations
systems, such as Maslow's hierarchy of human needs~\cite{maslow} (e.g., self-actualisation, security
respect of others). For each of these motivations, different roles that participate may have different {\it core motivations}, such as:

\begin{enumerate}
\item Benefit to contributor
\item Benefit to moderator
\item Benefit to system operator/host (e.g, Google, Facebook, Amazon)
\item To affiliates of the host (e.g., Amazon affiliates, YouTube partners)
\item To society as a whole
\item To a contributor's social network
\end{enumerate}

In addition to relationships to Maslow's hierarchy of needs, which drive the motivations,
we suggest that there are also resulting actions from each motivation. Thus, it should be
possible to follow basic human needs to motivations in social machines, through to the
actual actions (and therefore the necessary software feature support) that are made by
participants in social machines. Our theory is that this path may enable designers of
social machines to satisfy human needs by implemented specific features into their social
machines, or to identify potential gaps in their serving of needs and motivations due
to lack of implemented features.

Motivation in specific social machines has been studied, including motivation of participants
in: Wikipedia~\cite{kuznetsov2006motivations}; Open Source Software~\cite{lakhani2003hackers};
Peer-to-peer filesharing~\cite{p2p}; Tagging~\cite{tagging}; and Virtual
Communities~\cite{ardichvili2003motivation,Moore:2007:UMM:1235000.1235035}.

[relate this work back to ours briefly.]

\subsubsection{Artefacts of Social Machines}

Our next observation was that there are a number of shared artefact types that can be classified within the context of social machines. Specifically that
there are ``intentions'' that motivate the use of social machines, ``actions'' that are directly performed by participants of social machines, and ``contributions``
that result from the use of social machines. We have taken these abstract concepts, and mapped them to a number of well-known social machines in order
to demonstrate that these concepts can be used to compare and contrast different machines.

[more here]

\subsubsection{Metrics of participants and usage}

Finally, we identified several constructs that are relevant for all social machines. First, there are construct that relate to ``analytics'' that can be measured at a point in time, and may change over time. We have documented these constructs in Table~\ref{table:constructs}. In the left column we list construct regarding the design affordances of the social machine. In the middle column we have a mixture of the result of how the design affordances actually were realised by the social computation, such as how users co-opted / incorporated / interpreted the constraints. Finally, in the right column we list analytical measures of social machines which are subject to change over time.

\begin{table*}[htbp]
\begin{center}
\begin{tabular}{|p{18cm}|}
\hline
{\bf Designed Affordances} \\
\hline
Domain specificity [1 - Specific / 5 - General] \\
Visibility and importance of a participants `reputation' [1 - Not important / 5- Important] \\
Explicit representation of participant reputation [1 - No Reputations / 5 - Reputations] \\
Participants can define new types of contributions [1 - Not possible / 5 - Complete freedom] \\
Degree of open source software used/created [1 - Proprietary / 5 - Entirely open source] \\
Degree of openness and availability of data created in the system (APIs/Dumps) [1 - Closed/unavailable data / 5 - Completely open data] \\
Variety of actions participants can perform [1 - Single Action / 5 - Large variety of actions] \\
\hline
{\bf Participatory Affordances} \\
\hline
Supports social interaction [1 - No social features / 5 - Large variety of social features] \\
How often a participant participates (times a day, etc) [1 - Many times a day / 5 Seldom] \\
Importance of Timely Participation [1 - Unimportant / 5 - Timely] \\
Generality of Audience [1 - Niche / 5 - General] \\
Participant anonymity [1 - No Anonymity / 5 - Complete Anonymity] \\
Participant autonomy (1 - No autonomy / 5 - Complete autonomy) \\
Quality of participant contributions [1 - Low quality / 5 - High quality] \\
Clear separation of roles/responsibilities among participants - [1 - Single role - Everyone is the same) / 5 - People assume different roles/responsibilities] \\
Extent of hierarchical organisation of roles [1 - No Hierarchy / 5 - Deep/Well-defined Hierarchy] \\
Variety of types of contributions [1 - Single Type / 5 - Large variety of types] \\
Participants are motivated by extrinsic award  [1 - Disagree / 5 - Agree] \\
Participants are intrinsically motivated [1 - Disagree / 5 - Agree] \\
Service/Platform derives benefit from participant participation [1 - Disagree /  5- Agree] \\
Participants directly benefit from participating [1 - Disagree / 5 - Agree] \\
Participants' friends/social network benefit from participation [1 - Disagree / 5 - Agree] \\
Participation is for the benefit of a specific person/group other than the participant or platform owner [1-Disagree / 5-Agree] \\
Participation is for the benefit of society/the world at large  [1 - Disagree / 5 - Agree] \\
Participation is done to `get something done' [1-Disagree/5-Agree] \\
Participation is done `for fun' [1-Disagree/5-Agree] \\
Participation is done to exchange knowledge [1-Disagree/5-Agree] \\
Participation is done to be social [1-Disagree/5-Agree] \\
Participation is done via mobile devices [1-Never / 5 - Often] \\
Participants' (geographic) location is used by the service [1-No / 5 - Very] \\
\hline
{\bf Analytics} \\
\hline
Popularity [1 - Unpopular / 5 - Very Popular] \\
Maturity [1 - New / 5 - Mature] \\
Ratio of passive to active participants [1 - Passive / 5 - Active] \\
Number of competitors [1 - Not Many / 5 - Many] \\
Extent of global geographic coverage [1 - Highly localised / 5 - Global]\\
\hline
\end{tabular}
\end{center}
\caption{Constructs of social machines, listing affordances that are designed into machines, affordances that have resulted from use, and analytical constructs that can be measured and change over time.} \label{table:constructs}
\end{table*}%

The result is what we call `participation' constructs. These constructs arise from use of a
social machine, and are not necessarily predictable; they include:

\begin{enumerate}
\item {\bf Roles emerging in the social machine}
   \newline In addition to designed and implemented roles such as {\it moderator} and {\it administrator}, emerging roles include motivations of the participant, such as {\it trolls}, {\it spammers}, {\it domain experts}, {\it social nexus}, and {\it re-blogger}.
\item {\bf Quality of contributions}
    \newline For any social machine, the quality of contributions will vary. While some
    issues with quality can be predicted and protected against (e.g., through
    multi-user result agreement on Mechanical Turk~\cite{ipeirotis2010quality}), some
    particular issues with quality cannot always be predicted ahead of time.
\item {\bf Content of contributions}
    \newline While a system operator may design their social machine to generate specific
    types of content, participants may subvert the system in order to generate other types,
    or to concentrate on areas that were unforeseen at the time of launch, but are now seen
    as useful to the operators. It is also possible that multiple social machines can be
    combined/linked by users to further expand the types of contributions they produce.
\item {\bf Global / cultural reach}
    \newline A social machine may naturally focus on a narrow geographic area, for example
    an Ushahidi-based social machine for local elections~\cite{meier2008crisis} or
    natural disasters~\cite{morrow2011independent}. Such machines are unlikely to be
    re-purposed by participants outside of that area. However, some machines have been
    created for a single cultural/global area, but have potential global relevance that was
    unforeseen early on. An example is the fall-off of English-language users of the
    social network ``Orkut,'' which gained significant usage elsewhere, particularly in
    Brazil.
\item {\bf Task reach --- what do people use it for?}
    \newline In addition to breaking cultural barriers, user have also shown that they can
    use social machines for tasks that were not set up by the system operators. For example
    the identification of ``Green Pea Galaxies'' using the GalaxyZoo social machine for
    astronomy~\cite{greenpea}. Users were not originally asked to identify these types, but
    a number of users identified them, and the software was modified to include them,
    resulting in their discovery as a new type of galaxy.
\item {\bf Active/Passive participation roles}
    \newline When it comes to participation in social machines, there are levels of
    participation that can be classified as being ``active'' and being ``passive.'' These
    roles may not be particularly designed into systems, but may be exhibited over time.
    Particularly noteworthy examples can be seen in news and link aggregation social
    machines such as Reddit and Digg, where users can both submit links and vote on links.
    The frequency of each of these activities can vary wildly, with a small
    minority of users controlling the links that rise to the top~\cite{digg}.
\end{enumerate}

@@

\begin{figure}
\begin{center}
\includegraphics[width=4cm]{img/dendrogram-elements.png}
\caption{Dendrogram of elements (social machines) from a repertory grid exercise of the top 20 (ranked by Alexa) social machines from our element set, against our consolidated constructs.} \label{dendrogram}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/dendrogram-constructs.png}
\caption{Dendrogram of constructs from a repertory grid exercise of the top 20 (ranked by Alexa) social machines from our element set, against our consolidated constructs.} \label{dendrogram}
\end{center}
\end{figure}

>>>>

@@

\subsection{Methodology} \label{meth}

@@

\subsection{The Polyarchical Relationship of Social Machines}

While determining a set of social machines, one clear distinction was between social machine frameworks, such as MediaWiki and Ushahidi, which enable
social machines to be created, and instantiations of those frameworks into social machines that have been created for a specific purpose, such as Wikipedia
and Ushahidi Haiti. However, through further investigation, we discovered that the distinction between framework and instantiations was not the only
differentiator. Specifically, that some social machines could also have sub-machines or communities within them, that are working to solve different problems,
within the context of the larger machine. For example, Wikipedia is an instantiation of the MediaWiki framework, but there are a large number of
communities within Wikipedia that work on domain-specific areas, such as those contributing in areas of science and technology\footnote{Visualisation of Wikipedia Science Communities: \url{http://www.olihb.com/WikiCommunities/}}. However, we must also note that wikipedia is primarily split into communities of language, with the English, German and French wikipedias having the largest number of articles. It is therefore possible to analyse wikipedia into a hierarchy, from MediaWiki, to the languages of Wikipedia, down to the domain-specific communities. We explored this idea further, to include other social machines, and links where social machines utilise other social machines in order to function. For example, the social machine used by the Obama '12 US Presidential campaign~\cite{obamakieron}, which relied on social machines such as Twitter and Facebook.

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=8.5cm]{img/polyarchy.pdf}
\caption{A polyarchy of Social Machines, illustrating the infrastructure and frameworks used by social machines, and machine-machine usage.} \label{polyarchy}
\end{center}
\end{figure}

@@

%\subsubsection{Multi-faceted motivation, and who-gets-what benefit}
%
%%We identified a hierarchy of constructs pertaining to the motivation people have for using social
%%machines with top level constructs of:
%%
%%\begin{enumerate}
%%\item Hedonistic
%%\item Financial
%%\item To be informed
%%\item To help others
%%\end{enumerate}
%%
%%The sub constructs are illustrated in Figure~\ref{motivation}, and we relate them to Maslow's hierarchy of needs~\cite{maslow}, (e.g., self-actualisation, security, respect of others), in various ways.
%%
%%\begin{figure*}[htb]
%%\begin{center}
%%\includegraphics[width=18cm]{img/motivation.pdf}
%%\caption{Motivation hierarchy of participants using Social Machines, related to Maslow's hierarchy of needs.} \label{motivation}
%%\end{center}
%%\end{figure*}
%
%The motivations of users can be identified and mapped to existing human needs and motivations
%systems, such as Maslow's hierarchy of human needs~\cite{maslow} (e.g., self-actualisation, security
%respect of others). For each of these motivations, different roles that participate may have different {\it core motivations}, such as:
%
%\begin{enumerate}
%\item Benefit to contributor
%\item Benefit to moderator
%\item Benefit to system operator/host (e.g, Google, Facebook, Amazon)
%\item To affiliates of the host (e.g., Amazon affiliates, YouTube partners)
%\item To society as a whole
%\item To a contributor's social network
%\end{enumerate}
%
%In addition to relationships to Maslow's hierarchy of needs, which drive the motivations,
%we suggest that there are also resulting actions from each motivation. Thus, it should be
%possible to follow basic human needs to motivations in social machines, through to the
%actual actions (and therefore the necessary software feature support) that are made by
%participants in social machines. Our theory is that this path may enable designers of
%social machines to satisfy human needs by implemented specific features into their social
%machines, or to identify potential gaps in their serving of needs and motivations due
%to lack of implemented features.
%
%Motivation in specific social machines has been studied, including motivation of participants
%in: Wikipedia~\cite{kuznetsov2006motivations}; Open Source Software~\cite{lakhani2003hackers};
%Peer-to-peer filesharing~\cite{p2p}; Tagging~\cite{tagging}; and Virtual
%Communities~\cite{ardichvili2003motivation,Moore:2007:UMM:1235000.1235035}.
%
%[relate this work back to ours briefly.]

%\subsubsection{Artefacts of Social Machines}
%
%Our next observation was that there are a number of shared artefact types that can be classified within the context of social machines. Specifically that
%there are ``intentions'' that motivate the use of social machines, ``actions'' that are directly performed by participants of social machines, and ``contributions``
%that result from the use of social machines. We have taken these abstract concepts, and mapped them to a number of well-known social machines in order
%to demonstrate that these concepts can be used to compare and contrast different machines.
%
%[more here]

%\subsubsection{Metrics of participants and usage}
%
%Finally, we identified several constructs that are relevant for all social machines. First, there are construct that relate to ``analytics'' that can be measured at a point in time, and may change over time. We have documented these constructs in Table~\ref{table:constructs}. In the left column we list construct regarding the design affordances of the social machine. In the middle column we have a mixture of the result of how the design affordances actually were realised by the social computation, such as how users co-opted / incorporated / interpreted the constraints. Finally, in the right column we list analytical measures of social machines which are subject to change over time.

\subsection{Constructs}

The four clusters of constructs generated by the process described in Section~\ref{meth}
are shown in full in Table \ref{table:constructs}, and are detailed below.

\begin{table}[htb]
\begin{center}
\begin{tabular}{|p{8cm}|}
\hline
{\bf Tasks} \\
\hline
Variety of types of contributions \\
Domain specificity \\
Participation is done to `get something done' \\
Participation is done `for fun' \\
Participation is done to exchange knowledge \\
Participation is done to be social \\
Participants are intrinsically motivated \\
Participants are motivated by extrinsic reward \\
\hline
{\bf Users and Roles} \\
\hline
Generality of Audience \\
Participant autonomy \\
Participant anonymity \\
Extent of hierarchical organisation of roles \\
Clear separation of roles/responsibilities among participants \\
\hline
{\bf Participation} \\
\hline
Participants' (geographic) location is used by the service \\
Participation is done via: \\
\hspace{1cm} mobile devices \\
\hspace{1cm} web browsers \\
\hspace{1cm} apps \\
\hspace{1cm} sensors/sensing (location sensing and wearable devices) \\
Participation is done in a mobile context \\
\hline
{\bf Computation, Output and Benefit} \\
\hline
Output solves a computational problem \\
Output requires human insight \\
Output is a subjective appraisal \\
Participants directly benefit from participating \\
Service/Platform derives benefit from participant participation \\
Participants'' friends/social network benefit from participation \\
Participation is for the benefit of a specific person/group other than the participant or platform owner \\
Participation is for the benefit of society/the world at large  \\
\hline
\end{tabular}
\end{center}
\caption{Consolidated constructs of social machines.} \label{table:constructs}
\end{table}

\subsubsection{Tasks}

{\bf The types of tasks undertaken by the social component and the resulting contributions:}Social machines, just as any computing system, are built with a given purpose in mind, and exhibit functionality by which this purpose can be achieved. Users and the emerging social network interact with the system and with each other to carry out specific tasks by which they contribute to the 'computation'. This includes the creation and refinement of digital artifacts by executing given actions, possibly in collaboration with other participants. For example, in GalaxyZoo\footnote{\url{http://www.galaxyzoo.org}} people classify galaxies with collective performance as good as professional astronomers, while in reCaptcha\footnote{\url{http://www.google.com/recaptcha}} they refine the data that describes digital manuscripts. Sometimes, systems look for innovative ways of getting people to engage with the systems, including game narratives, like in EteRNA\footnote{\url{http://eterna.cmu.edu}}, in which  participants design RNA  that is evaluated automatically using simulators and then refined via human effort, or monetary rewards, like in Amazon's Mechanical Turk, a platform in which users undertake routine (micro) tasks such as object classification, audio transcription and text translation.

\subsubsection{Users and Roles}

{\bf The roles of the participants and the social network they form}

\subsubsection{Participation (How users interact with the Social Machine)}

\subsubsection{Computation, Output and Benefit}

%The result is what we call `participation' constructs. These constructs arise from use of a
%social machine, and are not necessarily predictable; they include:
%
%\begin{enumerate}
%\item {\bf Roles emerging in the social machine}
%   \newline In addition to designed and implemented roles such as {\it moderator} and {\it administrator}, emerging roles include motivations of the participant, such as {\it trolls}, {\it spammers}, {\it domain experts}, {\it social nexus}, and {\it re-blogger}.
%\item {\bf Quality of contributions}
%    \newline For any social machine, the quality of contributions will vary. While some
%    issues with quality can be predicted and protected against (e.g., through
%    multi-user result agreement on Mechanical Turk~\cite{ipeirotis2010quality}), some
%    particular issues with quality cannot always be predicted ahead of time.
%\item {\bf Content of contributions}
%    \newline While a system operator may design their social machine to generate specific
%    types of content, participants may subvert the system in order to generate other types,
%    or to concentrate on areas that were unforeseen at the time of launch, but are now seen
%    as useful to the operators. It is also possible that multiple social machines can be
%    combined/linked by users to further expand the types of contributions they produce.
%\item {\bf Global / cultural reach}
%    \newline A social machine may naturally focus on a narrow geographic area, for example
%    an Ushahidi-based social machine for local elections~\cite{meier2008crisis} or
%    natural disasters~\cite{morrow2011independent}. Such machines are unlikely to be
%    re-purposed by participants outside of that area. However, some machines have been
%    created for a single cultural/global area, but have potential global relevance that was
%    unforeseen early on. An example is the fall-off of English-language users of the
%    social network ``Orkut,'' which gained significant usage elsewhere, particularly in
%    Brazil.
%\item {\bf Task reach --- what do people use it for?}
%    \newline In addition to breaking cultural barriers, user have also shown that they can
%    use social machines for tasks that were not set up by the system operators. For example
%    the identification of ``Green Pea Galaxies'' using the GalaxyZoo social machine for
%    astronomy~\cite{greenpea}. Users were not originally asked to identify these types, but
%    a number of users identified them, and the software was modified to include them,
%    resulting in their discovery as a new type of galaxy.
%\item {\bf Active/Passive participation roles}
%    \newline When it comes to participation in social machines, there are levels of
%    participation that can be classified as being ``active'' and being ``passive.'' These
%    roles may not be particularly designed into systems, but may be exhibited over time.
%    Particularly noteworthy examples can be seen in news and link aggregation social
%    machines such as Reddit and Digg, where users can both submit links and vote on links.
%    The frequency of each of these activities can vary wildly, with a small
%    minority of users controlling the links that rise to the top~\cite{digg}.
%\end{enumerate}

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=4cm]{img/dendrogram-elements.png}
\caption{Dendrogram of elements (social machines) from a repertory grid exercise of the top 20 (ranked by Alexa) social machines from our element set, against our consolidated constructs.} \label{dendrogram}
\end{center}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=8cm]{img/dendrogram-constructs.png}
\caption{Dendrogram of constructs from a repertory grid exercise of the top 20 (ranked by Alexa) social machines from our element set, against our consolidated constructs.} \label{dendrogram}
\end{center}
\end{figure}
